
---
title: "Overlap of cell-type-specific eQTLs between Yazar_2022 and CSeQTL"
author: "Si Liu and Wei Sun"
date: "`r Sys.Date()`"
output: 
  html_document:
    theme: journal
    highlight: tango
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: true
      smooth_scroll: false
    number_sections: false
  df_print: paged
params:
  cs_q_cutoff: 5e-3
  fold_cutoff: 1
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(width = 80)
```

This code checks the overlap between the eGenes after combining cell types from Yazar et al. to larger categories, e.g., CD4T or CD8T. 

# Setup


## Load R libraries

```{r load_libraries, warning = FALSE, message = FALSE}
library(data.table)
library(readxl)
library(stringr)
library(ggplot2)
library(ggpubr)
library(dplyr)
library(reshape2)
theme_set(theme_classic())

Other_method = "Yazar_2022"

get_pi0 <- function(pval){
  mean(pval==1, na.rm = TRUE) + 
  2*sum((pval > 0.5) & (pval < 1), na.rm = TRUE) / sum(!is.na(pval))
}

len_unique <- function(x){
  return(length(unique(x)))
}

# params = list(cs_q_cutoff=5e-3, fold_cutoff=1)
cs_q_cutoff = params$cs_q_cutoff
fold_cutoff = params$fold_cutoff

cat("cs_q_cutoff: ",  cs_q_cutoff,  "\n")
cat("fold_cutoff: ",  fold_cutoff,  "\n")

config = sprintf("cs_q_%.0e_fold_%.1f", 
                 cs_q_cutoff, fold_cutoff)
config
```

## Data directory and file names
```{r}
CSeQTL_dir     = "../../CSeQTL/data2share/"
GTEx_eqtl_file = "GTExBlood_eqtls.rds"

Yazar_2022_dir = "../../CSeQTL/Yazar_2022/"

Yazar_2022_kept_folder = "Yazar_eSNP1_50kb/"
Yazar_2022_genes       = "Yazar_gene_considered.rds"
```

# Get overlap

## Load the genes considered for eQTL mapping in each cell type by Yazar et al. 

```{r}
Yazar_gene_list = readRDS(paste0(Yazar_2022_dir, Yazar_2022_genes))
length(Yazar_gene_list)

sapply(Yazar_gene_list, length)
sapply(Yazar_gene_list, len_unique)
```

## Combine cell types
```{r}
ct_groups = list(B = c("B_IN", "B_Mem"), 
                 CD4T = c("CD4_NC", "CD4_ET"),
                 CD8T = c("CD8_NC", "CD8_ET", "CD8_S100B"),
                 Monocyte = c("Mono_C", "Mono_NC"),
                 NK = c("NK_R", "NK"))

Yazar_gene_grp = list()

for(ct1 in names(ct_groups)){
  Yazar_gene_grp[[ct1]] = unique(unlist(Yazar_gene_list[ct_groups[[ct1]]]))
}

sapply(Yazar_gene_grp, length)
sapply(Yazar_gene_grp, len_unique)

```

## Load the genes considered under each cell type by CSeQTL

```{r}
GTExBlood <- readRDS(paste0(CSeQTL_dir, GTEx_eqtl_file))
lapply(GTExBlood, dim)

cs_result = GTExBlood$eqtl[GTExBlood$eqtl$trim==TRUE & 
                             GTExBlood$eqtl$perm==FALSE & 
                             GTExBlood$eqtl$MODEL=="cseqtl" & 
                             (!is.na(GTExBlood$eqtl$PVAL)), ]
dim(cs_result)
cs_result[1:2,]
table(cs_result$CELLTYPE)

cs_result = cs_result[-which(cs_result$CELLTYPE %in% 
                               c("Bulk", "OTHER")),]
dim(cs_result)
cs_result[1:2,]
table(cs_result$CELLTYPE)

# define a function to remove version number from gene id
rm_version <- function(id_w_version){
   gsub("\\..*", "", id_w_version)
}

cs_result$ENSG_version = cs_result$ENSG
cs_result$ENSG = rm_version(cs_result$ENSG_version)

ww2 = abs(log10(cs_result$ETA)) >= log10(fold_cutoff)
ww2 = ww2 & cs_result$qvalue < cs_q_cutoff

cs_result$eQTL = rep("NO", nrow(cs_result))
cs_result$eQTL[ww2] = "YES"

CSeQTL_gene_list = tapply(cs_result$ENSG, 
                          cs_result$CELLTYPE, function(v){v})

sapply(CSeQTL_gene_list, length)
```


## Cell type names

"cs" stands for CSeQTL and "ot" stands for other study, which is Yazar et al. 2022 here. 

```{r}
cs_cell_types = c("B", "CD4T", "CD8T", "Monocyte", "NK")
ot_cell_types = c("B", "CD4T", "CD8T", "Monocyte", "NK")
```


## Get the union of all the eGenes from CSeQTL

```{r}
table(cs_result$eQTL, cs_result$CELLTYPE, useNA = "ifany")

union_cs_eGenes = unique(cs_result$ENSG[cs_result$eQTL=="YES"])
length(union_cs_eGenes)
union_cs_eGenes[1:5]
```

## Compare results form CSeQTL and Yazar et al.

To avoid unbalanced number of eQTLs across cell types, keep top 300 eGenes for CSeQTL study and top 500 eGenes for Yazar et al.

```{r}

mat_cs_vs_ot= matrix(NA, nrow=length(cs_cell_types), 
                      ncol=length(ot_cell_types))
rownames(mat_cs_vs_ot) = cs_cell_types
colnames(mat_cs_vs_ot) = ot_cell_types

prop_mat_ot      = mat_cs_vs_ot
prop_mat_cs      = mat_cs_vs_ot
prop_mat_union   = mat_cs_vs_ot
ratio_obs_vs_exp = mat_cs_vs_ot
relative_ratios  = mat_cs_vs_ot
fisher_mat       = mat_cs_vs_ot
pi1_hat_mat_ot   = mat_cs_vs_ot
n_eGenes_ot      = mat_cs_vs_ot

pdf(width = 12, height = 6, 
    sprintf("figures/%s_grouped_eGenes_CSeQTL_pval_hist.pdf", 
            Other_method))

for (i in 1:length(ot_cell_types)){
  
  ot_ct    = ot_cell_types[i]
  ct_grp_i = ct_groups[[ot_ct]]
  
  ot_result_i = NULL
  
  for(ot_ct_j in ct_grp_i){
    ot_ct_str = sub(" ", "_", ot_ct_j)
    ot_eGenes_i_path = paste0(Yazar_2022_dir, Yazar_2022_kept_folder, 
                              "Yazar_eSNP1_50kb_", ot_ct_str, ".csv")
    
    ot_ct_result = read.csv(ot_eGenes_i_path, header = TRUE)
    ot_result_i  = rbind(ot_result_i, ot_ct_result)
  }

  dim(ot_result_i)
  ot_result_i[1:5,]
  table(ot_result_i$Cell.type)
  
  ot_eGenes_df = tapply(ot_result_i$qvalue, 
                       ot_result_i$Gene.Ensembl.ID, min)
  
  ot_eGenes_df = data.frame(Gene.Ensembl.ID = names(ot_eGenes_df), 
                            qvalue=ot_eGenes_df)
  ot_eGenes_df = ot_eGenes_df[order(ot_eGenes_df$qvalue),]
  summary(ot_eGenes_df$qvalue)
  
  cat(ot_ct," number of eGenes: ", nrow(ot_eGenes_df), "\n")

  ot_eGenes_i = ot_eGenes_df$Gene.Ensembl.ID[1:min(500,nrow(ot_eGenes_df))]
  
  # genes that are included in the eQTL mapping by Yazar et al.
  ot_ct_considered  = Yazar_gene_grp[[ot_ct]]

  # adjust the union of CSeQTL eGenes
  union_cs_eGenes_i = intersect(union_cs_eGenes, ot_ct_considered)
  
  # calculate base_union_ratio:
  # among the CSeQTL eGenes across all the cell types, what proportion 
  # overlaps with the eGenes in the i-th cell type by the other study

  base_union_ratio = length(intersect(ot_eGenes_i, union_cs_eGenes_i))
  base_union_ratio = base_union_ratio / length(union_cs_eGenes_i)

  p_ct_list = list()
  
  
  for (j in 1:length(cs_cell_types)){
    
    cs_ct = cs_cell_types[j]
    cs_ct_result = cs_result[cs_result$CELLTYPE == cs_ct,]
    cs_ct_result = cs_ct_result[order(cs_ct_result$qvalue),]
    cs_eGenes_j  = cs_ct_result$ENSG[cs_ct_result$eQTL=="YES"]
    
    if(length(cs_eGenes_j) > 300){ cs_eGenes_j = cs_eGenes_j[1:300] }

    # genes that are included in the eQTL mapping
    cs_ct_considered = CSeQTL_gene_list[[cs_ct]]
    
    overlap_considered = intersect(ot_ct_considered, cs_ct_considered)
    
    # filter the eGenes from Yazar 2022 and CSeQTLto keep those 
    # in the overlap of 'overlap_considered'
    
    ot_ij_eGenes = intersect(ot_eGenes_i, overlap_considered)
    cs_ij_eGenes = intersect(cs_eGenes_j, overlap_considered)
    
    n_eGenes_ot[j,i] = length(ot_ij_eGenes)
    
    eGenes_overlap = intersect(ot_ij_eGenes, cs_ij_eGenes)

    prop_mat_ot[j,i] = length(eGenes_overlap)/length(ot_ij_eGenes)
    prop_mat_cs[j,i] = length(eGenes_overlap)/length(cs_ij_eGenes)

    union_eGenes = union(ot_ij_eGenes, cs_ij_eGenes)
    prop_mat_union[j,i] = length(eGenes_overlap)/length(union_eGenes)
    
    # positive/negative by other method and CSeQTL.
    pp = length(eGenes_overlap)
    pn = length(setdiff(ot_ij_eGenes, eGenes_overlap))
    np = length(setdiff(cs_ij_eGenes, eGenes_overlap))
    nn = length(setdiff(overlap_considered, union_eGenes))
    
    stopifnot(pp+pn+np+nn == length(overlap_considered))
    
    # ratio between observed and expected
    ratio_obs_vs_exp[j,i] = (pp*(pp+pn+np+nn))/((pp+pn)*(pp+np))
    
    # relative ratio, compare the true discovery rate across cell types.
    TDR_cs_ij = pp/(pp + np)
    relative_ratios[j,i] = TDR_cs_ij/base_union_ratio

    # Fisher's exact test
    mat_test = matrix(c(pp, np, pn, nn), nrow = 2)
    rownames(mat_test) = c("Yazar eGenes", "Yazar Non-eGenes")
    colnames(mat_test) = c("CSeQTL eGenes", "CSeQTL Non-eGenes")

    cat("Other:", ot_ct, "CSeQTL:", cs_ct, "\n")
    print(mat_test)
    cat("\n")
    
    fisher_mat[j,i] = fisher.test(mat_test, 
                                  alternative = "greater")$p.value
    
    # get CSeQTL pvalue for the eGenes by the other method
    ot_eGenes_CSeQTL_pvalues = 
    cs_ct_result[match(ot_ij_eGenes, cs_ct_result$ENSG),]$permutation_PVAL
    
    # Use p-value distribution to estimate the proportion of null
    pi0_hat = get_pi0(ot_eGenes_CSeQTL_pvalues)
    pi0_hat = min(1, pi0_hat)
    pi1_hat_mat_ot[j,i] = 1 - pi0_hat

    # p-value histogram
    df_ct_plot_pvalues = data.frame(pvalue = ot_eGenes_CSeQTL_pvalues)
    
    p_ct_list[[j]] = ggplot(df_ct_plot_pvalues, aes(x=pvalue)) +
      geom_histogram(color="darkblue", fill="lightblue",
                     breaks=seq(0,1,by=0.02)) + 
                     ggtitle(paste0(Other_method, ": ", ot_ct, 
                                    "\nCSeQTL: ", cs_ct, 
                                    "\npi_0: ", signif(pi0_hat, 2)))
  }
  
  print(ggarrange(plotlist=p_ct_list, nrow = 2, ncol = 4))
  
}

dev.off()

n_eGenes_ot
```

## Plots
```{r fig.width=4.2, fig.height=3}
theme1 = theme(axis.text.x = element_text(angle = 60, vjust = 1, 
                                   size = 11, hjust = 1),
               axis.text.y = element_text(size = 11))

# among eGenes of other method, what proportion was recovered by CSeQTL
melted_prop_mat_ot = reshape2::melt(prop_mat_ot,
                          varnames=c("CSeQTL", Other_method))
names(melted_prop_mat_ot)[1:2] = c("CSeQTL", Other_method)

round(prop_mat_ot,2)

ggplot(data = melted_prop_mat_ot, 
       aes_string(x="CSeQTL", y=Other_method, fill="value")) + 
  geom_tile() + theme_classic() + 
  scale_fill_gradient(low = "white", high = "red") + 
  ggtitle(paste("Prop. overlap among", 
                Other_method, "eGenes")) + theme1

melted_prop_mat_cs = reshape2::melt(prop_mat_cs,
                          varnames=c("CSeQTL", Other_method))

round(prop_mat_cs, 2)
ggplot(data = melted_prop_mat_cs, 
       aes_string(x="CSeQTL", y=Other_method, fill="value")) + 
  geom_tile() + theme_classic() + 
  scale_fill_gradient(low = "white", high = "red") +  
  ggtitle("Prop. overlap among CSeQTL eGenes") + theme1


melted_prop_mat_union = reshape2::melt(prop_mat_union, 
                              varnames=c("CSeQTL", Other_method))

round(prop_mat_union, 2)
ggplot(data = melted_prop_mat_union, 
       aes_string(x="CSeQTL", y=Other_method, fill="value")) + 
  geom_tile() + theme_classic() + 
  scale_fill_gradient(low = "white", high = "red")+ 
  ggtitle("Prop. overlap among union of eGenes") + theme1


melted_ratio = reshape2::melt(ratio_obs_vs_exp,
                    varnames=c("CSeQTL", Other_method))

round(ratio_obs_vs_exp, 2)
ggplot(data = melted_ratio, 
       aes_string(x="CSeQTL", y=Other_method, fill="value")) + 
  geom_tile() + theme_classic() + 
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
                       midpoint = 1) + 
  ggtitle("ratio of observed/expected overlap") + theme1


melted_relative_ratios = reshape2::melt(relative_ratios,
                            varnames=c("CSeQTL", Other_method))

round(relative_ratios, 2)
ggplot(data = melted_relative_ratios[!is.na(melted_relative_ratios$value),], 
       aes_string(x="CSeQTL", y=Other_method, fill="value")) + 
  geom_tile() + theme_classic() + 
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
                       midpoint = 1) + 
  ggtitle("relative ratio") + theme1

fisher_mat_cut = fisher_mat
fisher_mat_cut[fisher_mat < (10^(-10))] = 10^(-10)

melted_fisher_mat = reshape2::melt(-log10(fisher_mat_cut),
                      varnames=c("CSeQTL", Other_method))

round(-log10(fisher_mat_cut), 2)
ggplot(data = melted_fisher_mat, 
       aes_string(x="CSeQTL", y=Other_method, fill="value")) + 
  geom_tile() + theme_classic() + 
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
                       midpoint = -log10(0.05), name="-log10(pvalue)") + 
  ggtitle("-log10(pvalue) for overlap") + theme1

melted_pi1 = reshape2::melt(pi1_hat_mat_ot, 
                varnames=c("CSeQTL", Other_method))

round(pi1_hat_mat_ot,3)
ggplot(data = melted_pi1, 
       aes_string(x="CSeQTL", y=Other_method, fill="value")) + 
  geom_tile() + theme_classic() + 
  scale_fill_gradient(low = "white", high = "red") +  
  ggtitle(paste("pi1 of", Other_method, "eGenes")) + 
  theme1
```


## Write out results

```{r}
write.csv(as.data.frame(prop_mat_cs), row.names = TRUE, 
          file = sprintf("results/step3b_%s_prop_overlap_among_CSeQTL.csv",
                         config))

write.csv(as.data.frame(ratio_obs_vs_exp), row.names = TRUE, 
          file = sprintf("results/step3b_%s_ratio_obs_vs_exp.csv",
                         config))

write.csv(as.data.frame(relative_ratios), row.names = TRUE, 
          file = sprintf("results/step3b_%s_relative_ratios.csv",
                         config))  

write.csv(as.data.frame(fisher_mat), row.names = TRUE, 
          file = sprintf("results/step3b_%s_overlap_fisher_pvalue.csv",
                         config))
```

# Session information
```{r}
gc()

sessionInfo()
```



