---
title: "Overlap between Bryois 2022 and CMC case by Cell Type"
author: "Si Liu and Wei Sun"
date: "`r Sys.Date()`"
output: 
  html_document:
    theme: journal
    highlight: tango
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: true
      smooth_scroll: false
    number_sections: false
  df_print: paged
params:
  cs_q_cutoff: 5e-3
  fold_cutoff: 1.5
  AF: SCZ
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(width = 80)
```

Check the overlap between the eGenes from Bryois et al. (2022) and eGenes identified by CSeQTL from Common Mind Consortium (CMC) Schizophrenia patients. 


# Setup

## Load R libraries
```{r load_libraries, warning = FALSE, message = FALSE}
library(data.table)
library(readxl)
library(stringr)
library(ggplot2)
library(ggpubr)
library(dplyr)
library(reshape2)

theme_set(theme_classic())

Other_method = "Bryois_2022"

get_pi0 <- function(pval){
  mean(pval==1, na.rm = TRUE) + 
  2*sum((pval > 0.5) & (pval < 1), na.rm = TRUE) / sum(!is.na(pval))
}

is.unq <- function(x) length(x)==length(unique(x))

# params = list(cs_q_cutoff=5e-3, fold_cutoff=1, AF="SCZ")
cs_q_cutoff = params$cs_q_cutoff # q value cutoff
fold_cutoff = params$fold_cutoff # fold change cutoff
AF = params$AF                   # Analysis Filter, SCZ vs. Control 

cat("cs_q_cutoff: ",  cs_q_cutoff,  "\n")
cat("fold_cutoff: ",  fold_cutoff,  "\n")
cat("AF: ",  AF,  "\n")

config = sprintf("%s_cs_q_%.0e_fold_%.1f", 
                 AF, cs_q_cutoff, fold_cutoff)
config
```

## Data directory and file names
```{r}
CSeQTL_dir  = "../../CSeQTL/data2share/"
CSeQTL_file = "CMC_eqtls.rds"

Bryois_2022_data_dir = "../../CSeQTL/Bryois_2022/"

Bryois_eGene_folder  = "Bryois_eGenes_50kb/"
Bryois_con_gene_file = "Bryois_gene_considered.rds"
```

## cell type names

```{r}

Bryois_cell_types = c("Astrocytes", "Excitatory neurons", 
                      "Inhibitory neurons",  "Microglia", 
                      "Oligodendrocytes", "OPCs / COPs", 
                      "Pericytes", "Endothelial cells")

```

## Identify eGenes from Bryois 2022

```{r}
eGenes_Bryois_list = list()
n_eGenes = list()

for (i in 1:length(Bryois_cell_types)){
  
  Bryois_ct = Bryois_cell_types[i]
  Bryois_ct_string = sub(" ", "_", str_squish(sub("/", "", Bryois_ct)))
  
  Bryois_ct_eGenes_path = paste0(Bryois_2022_data_dir, 
                                 Bryois_eGene_folder, 
                                 "Bryois_eGenes_50kb_", 
                                 Bryois_ct_string, ".csv")
  
  df_Bryois_ct  = read.csv(Bryois_ct_eGenes_path, header = TRUE)
  
  cat(i, Bryois_ct, nrow(df_Bryois_ct), "\n")
  print(summary(abs(df_Bryois_ct$beta)))
  
  table(df_Bryois_ct$adj_p == 0)
  ggplot(df_Bryois_ct, aes(x=beta, y=-log10(adj_p))) + 
    geom_point(size = 1, alpha = 0.5) +  
    geom_vline(xintercept=0.3, col="red") + 
    geom_vline(xintercept=-0.3, col="red")
  
  w2kp = which(abs(df_Bryois_ct$beta) > 0.3)
  df_Bryois_ct = df_Bryois_ct[w2kp,]
  cat(i, Bryois_ct, "after beta filtering:", 
      nrow(df_Bryois_ct), "\n\n")

  df_Bryois_ct = df_Bryois_ct[order(df_Bryois_ct$adj_p),]
  n_eGenes[[Bryois_ct_string]] = nrow(df_Bryois_ct)

  if(nrow(df_Bryois_ct) > 500){
    df_Bryois_ct = df_Bryois_ct[1:500,]
  }
  
  eGenes_Bryois_list[[Bryois_ct_string]] = df_Bryois_ct$ensembl
  
}

sapply(eGenes_Bryois_list, length)
sapply(eGenes_Bryois_list, function(v){length(unique(v))})

```

## Cell type selection

"Pericytes" and "Endothelial" have fewer eGenes, so we skip it. 

```{r}

Bryois_cell_types = c("Astrocytes", "Excitatory neurons", 
                      "Inhibitory neurons",  "Microglia", 
                      "Oligodendrocytes", "OPCs / COPs")

CSeQTL_cell_types = c("Astro", "Exc", "Inh", "Micro", "Oligo", "OPC")
```

# Get overlap

## Step 1: load the genes considered under each cell type by Bryois_2022

```{r}
gene_con_Bryois_list = readRDS(paste0(Bryois_2022_data_dir, 
                                      Bryois_con_gene_file))

sapply(gene_con_Bryois_list, length)
```

## Step 2: load the genes considered for each cell type by CSeQTL, and extract the eGenes.

```{r}
CSeQTL <- readRDS(paste0(CSeQTL_dir, CSeQTL_file))
length(CSeQTL)
names(CSeQTL)

lapply(CSeQTL, dim)

CSeQTL = CSeQTL[["eqtl"]]
table(CSeQTL$AF)

CSeQTL = CSeQTL[which(CSeQTL$AF==AF),]
dim(CSeQTL)
head(CSeQTL)
colSums(!is.na(CSeQTL))

length(unique(CSeQTL$ENSG))
length(unique(CSeQTL$GENE))
length(unique(CSeQTL$SNP))

table(CSeQTL$CELLTYPE, useNA = "ifany")

CSeQTL$ENSG_version = CSeQTL$ENSG
CSeQTL$ENSG = gsub("\\..*","", CSeQTL$ENSG_version)

CSeQTL = CSeQTL[which((CSeQTL$trim==TRUE) & 
                        (CSeQTL$perm==FALSE) & 
                        (CSeQTL$MODEL=="cseqtl") & 
                        (!is.na(CSeQTL$PVAL))), ]

ww2 = abs(log10(CSeQTL$ETA)) >= log10(fold_cutoff)
ww2 = ww2 & CSeQTL$qvalue < cs_q_cutoff

CSeQTL$eQTL = rep("NO", nrow(CSeQTL))
CSeQTL$eQTL[ww2] = "YES"

n_con_CSeQTL_list    = list()
gene_con_CSeQTL_list = list()

n_eGenes_CSeQTL_list = list()
eGenes_CSeQTL_list   = list()

for (j in 1:length(CSeQTL_cell_types)){
  
  cct = CSeQTL_cell_types[j]
  
  cct_CSeQTL = CSeQTL[which(CSeQTL$CELLTYPE == cct), ]
  
  cct_considered_CSeQTL = unique(cct_CSeQTL$ENSG)
  
  n_con_CSeQTL_list[[cct]] = length(cct_considered_CSeQTL)
  gene_con_CSeQTL_list[[cct]] = cct_considered_CSeQTL
  
  cct_CSeQTL_eGenes = 
    unique(cct_CSeQTL$ENSG[which(cct_CSeQTL$eQTL=="YES")])

  n_eGenes_CSeQTL_list[[cct]] = length(cct_CSeQTL_eGenes)
  eGenes_CSeQTL_list[[cct]] = cct_CSeQTL_eGenes

}

unlist(n_con_CSeQTL_list)
unlist(n_eGenes_CSeQTL_list)
```


## Further filter eQTLs

To avoid unbalanced number of eQTLs across cell types, keep top 300 eGenes if there are more than 300 eGenes 

```{r}

for (j in 1:length(CSeQTL_cell_types)){
  cct = CSeQTL_cell_types[j]
  cct_CSeQTL = CSeQTL[which(CSeQTL$CELLTYPE == cct), ]
  
  if(n_eGenes_CSeQTL_list[[cct]] > 300){
    cct_CSeQTL = cct_CSeQTL[order(cct_CSeQTL$qvalue),]
    eGenes_CSeQTL_list[[cct]] = cct_CSeQTL$ENSG[1:300]
  }
}
sapply(eGenes_CSeQTL_list, length)

for (j in 1:length(CSeQTL_cell_types)){
  cct = CSeQTL_cell_types[j]
  int_j = intersect(eGenes_CSeQTL_list[[cct]], eGenes_CSeQTL_list[["Exc"]])
  cat(cct, length(int_j), "\n")
}

```

## Step 3: get the union of all eGenes from CSeQTL
```{r}
union_CSeQTL_eGenes = unique(c(eGenes_CSeQTL_list[["Astro"]], 
                               eGenes_CSeQTL_list[["Exc"]], 
                               eGenes_CSeQTL_list[["Inh"]], 
                               eGenes_CSeQTL_list[["Micro"]],
                               eGenes_CSeQTL_list[["Oligo"]], 
                               eGenes_CSeQTL_list[["OPC"]]))
length(union_CSeQTL_eGenes)
```

## Step 4: check overlap

```{r}
# "cs" stands for CSeQTL and "ot" stands for other study
# in this case, th other study is  Bryois 2022. 

mat_cs_vs_ot= matrix(NA, nrow=length(CSeQTL_cell_types), 
                      ncol=length(Bryois_cell_types))

rownames(mat_cs_vs_ot) = CSeQTL_cell_types
colnames(mat_cs_vs_ot) = Bryois_cell_types

prop_mat_ot      = mat_cs_vs_ot
prop_mat_cs      = mat_cs_vs_ot
prop_mat_union   = mat_cs_vs_ot
ratio_obs_vs_exp = mat_cs_vs_ot
relative_ratios  = mat_cs_vs_ot
fisher_mat       = mat_cs_vs_ot
pi1_hat_mat_ot   = mat_cs_vs_ot

eGenes_overlap_list = list()

pdf(width = 12, height = 6, 
    sprintf("figures/step4_Bryois_eGenes_CSeQTL_CMC_%s_pval_hist.pdf", AF))

for (i in 1:length(Bryois_cell_types)){
  
  Bryois_ct = Bryois_cell_types[i]
  Bryois_ct_string = sub(" ", "_", str_squish(sub("/", "", Bryois_ct)))
  
  Bryois_ct_eGenes     = eGenes_Bryois_list[[Bryois_ct_string]]
  Bryois_ct_considered = gene_con_Bryois_list[[Bryois_ct_string]]

  # filter CSeQTL eGenes to keep those considered by Bryois et al.
  union_CSeQTL_eGenes_i = intersect(union_CSeQTL_eGenes, 
                                    Bryois_ct_considered)
  
  stopifnot(is.unq(union_CSeQTL_eGenes_i))

  base_union_ratio = intersect(Bryois_ct_eGenes, 
                               union_CSeQTL_eGenes_i)
  base_union_ratio = length(base_union_ratio)/length(union_CSeQTL_eGenes_i)

  print(paste0("Bryois cell type: ", Bryois_ct_string))
  print(paste0("# of Bryois eGenes: ", length(Bryois_ct_eGenes)))
  print(paste0("length of union_CSeQTL_eGenes: ", 
               length(union_CSeQTL_eGenes_i)))
      
  p_ct_list = list()
  
  for (j in 1:length(CSeQTL_cell_types)){
    
    CSeQTL_ct = CSeQTL_cell_types[j]
    CSeQTL_ct_eGenes = eGenes_CSeQTL_list[[CSeQTL_ct]]
    
    # get the overlap of genes considered by two studies
    CSeQTL_ct_considered = gene_con_CSeQTL_list[[CSeQTL_ct]]
    
    overlap_considered = intersect(Bryois_ct_considered, 
                                   CSeQTL_ct_considered)
    
    # Filter the eGenes to keep those in 'overlap_considered'
    Bryois_ct_eGenes_kept = intersect(Bryois_ct_eGenes, 
                                      overlap_considered)

    CSeQTL_ct_eGenes_kept = intersect(CSeQTL_ct_eGenes, 
                                      overlap_considered)
    
    if(j == i){
      eGenes_overlap_list[[CSeQTL_ct]] = list(Bryois=Bryois_ct_eGenes_kept, 
                                         CSeQTL=CSeQTL_ct_eGenes_kept)
    }

    # eGenes overlap
    eGenes_overlap = intersect(Bryois_ct_eGenes_kept, 
                               CSeQTL_ct_eGenes_kept)
    n_overlap = length(eGenes_overlap)
    
    prop_mat_ot[j,i] = n_overlap/length(Bryois_ct_eGenes_kept)
    prop_mat_cs[j,i] = n_overlap/length(CSeQTL_ct_eGenes_kept)
    
    union_eGenes = union(Bryois_ct_eGenes_kept, 
                         CSeQTL_ct_eGenes_kept)
    prop_mat_union[j,i] = n_overlap/length(union_eGenes)

    pp = n_overlap
    pn = length(setdiff(Bryois_ct_eGenes_kept, eGenes_overlap))
    np = length(setdiff(CSeQTL_ct_eGenes_kept, eGenes_overlap))
    nn = length(setdiff(overlap_considered, union_eGenes))

    stopifnot(pp+pn+np+nn == length(overlap_considered))

    # ratio between observed and expected
    ratio_obs_vs_exp[j,i] = (pp*(pp+pn+np+nn))/((pp+pn)*(pp+np))
    
    # relative ratio, compare true discovery rate across cell types.
    TDR_cs_ij = pp/(pp + np)
    relative_ratios[j,i] = TDR_cs_ij/base_union_ratio
    
    # Fisher's exact test
    mat_test = matrix(c(pp, np, pn, nn), nrow = 2)
    rownames(mat_test) = c("Bryois eGenes", "Bryois Non-eGenes")
    colnames(mat_test) = c("CSeQTL eGenes", "CSeQTL Non-eGenes")
    
    cat("Other:", Bryois_ct, "CSeQTL:", CSeQTL_ct, "\n")
    print(mat_test)
    cat("\n")
    
    fisher_mat[j,i] = fisher.test(mat_test, 
                                  alternative = "greater")$p.value
    
    # get CSeQTL p-values for the eGenes by the other method
    CSeQTL_ct_res = CSeQTL[which(CSeQTL$CELLTYPE == CSeQTL_ct), ]

    ot_eGenes_CSeQTL_pvalues = 
      CSeQTL_ct_res$permutation_PVAL[match(Bryois_ct_eGenes_kept, 
                         CSeQTL_ct_res$ENSG)]
      
    pi0_hat = get_pi0(ot_eGenes_CSeQTL_pvalues)
    pi0_hat = min(1, pi0_hat)
    pi1_hat_mat_ot[j,i] = 1 - pi0_hat
    
    df_ct_plot_pvalues = data.frame(pvalue = ot_eGenes_CSeQTL_pvalues)
    
    p_ct_list[[j]] = ggplot(df_ct_plot_pvalues, aes(x=pvalue)) +
      geom_histogram(color="darkblue", fill="lightblue",
                     breaks=seq(0,1,by=0.02)) + 
                     ggtitle(paste0(Other_method, ": ", Bryois_ct, 
                                    "\nCSeQTL: ", CSeQTL_ct, 
                                    "\npi_0: ", signif(pi0_hat, 2)))
  }
  
  print(ggarrange(plotlist=p_ct_list, nrow = 2, ncol = 4))
  
}

dev.off()

sapply(eGenes_overlap_list, function(x){sapply(x, length)})
```

## Plots
```{r fig.width=4.8, fig.height=3.6}
theme1 = theme(axis.text.x = element_text(angle = 60, vjust = 1, 
                                   size = 11, hjust = 1),
               axis.text.y = element_text(size = 11))

# among eGenes of other method, what proportion was recovered by CSeQTL
melted_prop_mat_ot = reshape2::melt(prop_mat_ot,
                          varnames=c("CSeQTL", Other_method))
names(melted_prop_mat_ot)[1:2] = c("CSeQTL", Other_method)

round(prop_mat_ot,2)

ggplot(data = melted_prop_mat_ot, 
       aes_string(x="CSeQTL", y=Other_method, fill="value")) + 
  geom_tile() + theme_classic() + 
  scale_fill_gradient(low = "white", high = "red") + 
  ggtitle(paste("Prop. overlap among", 
                Other_method, "eGenes")) + theme1

melted_prop_mat_cs = reshape2::melt(prop_mat_cs,
                          varnames=c("CSeQTL", Other_method))

round(prop_mat_cs, 2)
ggplot(data = melted_prop_mat_cs, 
       aes_string(x="CSeQTL", y=Other_method, fill="value")) + 
  geom_tile() + theme_classic() + 
  scale_fill_gradient(low = "white", high = "red") +  
  ggtitle("Prop. overlap among CSeQTL eGenes") + theme1

melted_prop_mat_union = reshape2::melt(prop_mat_union, 
                              varnames=c("CSeQTL", Other_method))

round(prop_mat_union, 2)
ggplot(data = melted_prop_mat_union, 
       aes_string(x="CSeQTL", y=Other_method, fill="value")) + 
  geom_tile() + theme_classic() + 
  scale_fill_gradient(low = "white", high = "red")+ 
  ggtitle("Prop. overlap among union of eGenes") + theme1


melted_ratio = reshape2::melt(ratio_obs_vs_exp,
                    varnames=c("CSeQTL", Other_method))

round(ratio_obs_vs_exp, 2)
ggplot(data = melted_ratio, 
       aes_string(x="CSeQTL", y=Other_method, fill="value")) + 
  geom_tile() + theme_classic() + 
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
                       midpoint = 1) + 
  ggtitle("ratio of observed/expected overlap") + theme1

melted_relative_ratios = reshape2::melt(relative_ratios,
                            varnames=c("CSeQTL", Other_method))

round(relative_ratios, 2)
ggplot(data = melted_relative_ratios[!is.na(melted_relative_ratios$value),], 
       aes_string(x="CSeQTL", y=Other_method, fill="value")) + 
  geom_tile() + theme_classic() + 
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
                       midpoint = 1) + 
  ggtitle("relative ratio") + theme1

fisher_mat_cut = fisher_mat
fisher_mat_cut[fisher_mat < (10^(-10))] = 10^(-10)

melted_fisher_mat = reshape2::melt(-log10(fisher_mat_cut),
                      varnames=c("CSeQTL", Other_method))

round(-log10(fisher_mat_cut), 2)
ggplot(data = melted_fisher_mat, 
       aes_string(x="CSeQTL", y=Other_method, fill="value")) + 
  geom_tile() + theme_classic() + 
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
                       midpoint = -log10(0.05), name="-log10(pvalue)") + 
  ggtitle("-log10(pvalue) for overlap") + theme1

melted_pi1 = reshape2::melt(pi1_hat_mat_ot, 
                varnames=c("CSeQTL", Other_method))

round(pi1_hat_mat_ot,3)
ggplot(data = melted_pi1, 
       aes_string(x="CSeQTL", y=Other_method, fill="value")) + 
  geom_tile() + theme_classic() + 
  scale_fill_gradient(low = "white", high = "red") +  
  ggtitle(paste("pi1 of", Other_method, "eGenes")) + 
  theme1
```


## Write out results

```{r}
write.csv(as.data.frame(prop_mat_cs), row.names = TRUE, 
          file = sprintf("results/step4_%s_prop_overlap_among_CSeQTL.csv",
                         config))

write.csv(as.data.frame(ratio_obs_vs_exp), row.names = TRUE, 
          file = sprintf("results/step4_%s_ratio_obs_vs_exp.csv",
                         config))

write.csv(as.data.frame(relative_ratios), row.names = TRUE, 
          file = sprintf("results/step4_%s_relative_ratios.csv",
                         config))  

write.csv(as.data.frame(fisher_mat), row.names = TRUE, 
          file = sprintf("results/step4_%s_overlap_fisher_pvalue.csv",
                         config))

saveRDS(eGenes_overlap_list, 
        file=sprintf("results/step4_%s_eGenes.rds", config))

```

# Session information
```{r}
gc()

sessionInfo()
```
